---
title: "Adult Income Study: Milestone 01"
author: "Jimmy Liu and Hannah McSorley"
date: "29/02/2020"
output:
  html_document:
    keep_md: true
---

# Adult Income Study
```{r load-library, echo = F, warning = F, message = F}

knitr::opts_chunk$set(echo=FALSE, warning=FALSE)

library(tidyverse)
library(here)
library(DT)
```

```{r load-data}
# read csv file
dat <- read.csv(here("data/adult.data"), header = F)

# rename columns
names(dat) <- c("age", "workclass", "fnlwgt", "education", "education-num", "martial_status", 
                "occupation", "relationship", "race", "sex", "capital-gain", "capital-loss", 
                "hours-per-week", "native-country", "label")
```

## Introduction 

The economic well-being of individuals is reliant on their income, where income is defined as the money an individual (or household) receives on a regular basis. In the United States, the Census Bureau uses income  (money received before expenses and deductions) to gauge the population's range of poverty, wealth, and financial security (United States Census Bureau, 2016). There are a variety of factors that can influence one's income, including socioeconomic drivers, education and vocation. This project examines some of the variables that are often related to income. 

## Description of Dataset

This project works with a dataset of adult incomes obtained from the University of California Irvine (UCI) [Machine Learning Repository](https://archive.ics.uci.edu/ml/datasets/adult). The data was donated by Ronny Kohavi and Barry Becker (Silicon Graphics) and was originally extracted by Barry Becker from the 1994 Census database and used for machine learning predictions of whether a person makes over $50,000 per year based on personal factors.

This 1994 income census dataset consists of multivariate categorical and integer data that describe socioeconomic and personal classifiers of adults across the USA. Each instance (32,561) is an individual whose annual income was grouped as either above or below $50,000. Table 1 shows an overview of the 15 attributes (variables), including whether each is categorical or integer and a brief interpretation of the variable.  

<br>
```{r, variable-table, echo = FALSE}
read_csv(here::here("data/variable_summary.csv"), col_names = TRUE) %>% 
  DT::datatable(caption = "description of adult income dataset variables")
```
<br>

A couple of assumptions were made about these data based on information on the Census website. It is assumed that "capital gains" indicate non-cash financial benefits (e.g., food stamps, health benefits, subsidized housing or transportation, employer contributions to retirement programs, medical and educational expenses, etc.), and that "capital losses" include non-cash expenses (such as depreciated value of assets). We are also assuming that "education number" indicates the number of years allotted to education.  

It is of note that these data are from 1994 census, and the income threshold of $50,000 held a different meaning for wealth than it holds today. As this dataset includes socioeconomic attributes, it's worth noting that US-born white males comprise the majority of the data instances. 

## Exploratory Data Analysis 

### Summary overview

```{r, summary}
summary(dat)
```

The summary overview provides a snapshot of the data spread and averages. From this, we see this dataset includes a disproportionate number of middle-age, white, US-born, private-sector employees. There appears to be a fairly even distribution of individuals across occupational sectors and the majority of individuals work approximately 40 hours per week.

## Research Questions

In this study, we will explore the relationships between personal attributes and quantitative income-related variables with the goal of identifying relationships and interesting patterns. Specifically, we will focus on addressing the following exploratory research questions: 

1. Is there an observable relationship between personal attributes data and income level?
2. Does the number of hours worked per week relate more to occupation, sex, race, age, or is there no clear relationship?
3. What is the relationship between education and hours worked per week (e.g. does a person work fewer hours if they have completed more schooling)?

## Plan of Action

The variables that effect income may be confounding and are unlikely to be direct, therefore these data may not be appropriate for linear regression analyses. We will focus on exploring the relationships variables and identifying relationships and patterns. 


## References

United States Census Bureau, 2016. Income and Poverty, 'about income'. https://www.census.gov/topics/income-poverty/income/about.html

University of California Irvine, Machine Learning Repository. https://archive.ics.uci.edu/ml/datasets/adult.